---
title: "Mapping the shifts: analysis of recreational cannabis policies from 2013-2024 in US states"

author: "Ariadne Rivera"

format: 
  html:
    toc: true

editor: visual

bibliography: references.bib
---

```{r}
#| label: load-packages 
#| echo: false 
#| warning: false

library(ggcorrplot) 
library(tidyverse) 
library(haven) 
library(Rtsne) 
library(factoextra) 
library(corrplot) 
library(here) 
library(MplusAutomation) 
library(ggpubr) 
library(psych) 
library(glue) 
library(gt) 
library(reshape2) 
library(cowplot) 
library(vcd) 
library(corrr)
library(stringr)
library(ggtext)
library(patchwork)
library(rcartocolor)
library(grid)
library(forcats)
  
```

# Background

Over the past two decades, rapid shifts have occurred in the legal landscape of cannabis. Many states have moved from a prohibitionist type of approach to legalization for medical use and towards the subsequent legalization of cannabis legalization for recreational use. As of November 2024, 39 states have permitted the use and commercialization of medical cannabis products, 9 states allowed the use low THC, high CBD products, and 24 states had legalized cannabis for recreational purposes [@ncls2024]. Consequently, approximately 80% of the population in the US resides in a county where they can access medical or adult-use cannabis through a retail outlet [@shah2024].

This quarto file includes the final code to create figures and results from the models presented in the thesis.

```{r}
#| label: load-dataset
#| echo: false
#| warning: false

df <- read.csv(here(
  "data/3_analytic/cannabis_lca_analytic_2013_2024_updated.csv"
))

#names(df)

#str(df)

df <- df %>%
  rename(ID = id)

state <- subset(df, select = c("ID", "styr", "stname", "stabbr", "stfips", "year"))
state <- state %>% arrange(styr)

```

## Descriptive

### Correlation plots of cannabis characteristics (continuous)

```{r}
#| label: listofcovs for corrplots
#| echo: false
#| warning: false

col_names <- c("rml", "rmlsl", "regag", "lcauth", "edib", "conc",
               "thclim", "rtl", "pblus", "rmlhc", "hmdl", "estlm", 
               "tax2", "prcng", "pkr", "warnr", "adgn", "adyth",
               "undrg2", #"drvpr2", 
               "drvpr3",  "opncn")

```

##### Correlation of RML chacatersitics among RML states

This figure shows the correlation between recreational cannabis policies characteristics between 2013 to 2024 among RML state-years only.

```{r corrplot, fig.width=8, fig.height=8, out.extra="width=0.9\\linewidth"}
#| label: corrplot rml states
#| echo: false 
#| warning: false

df_rml <- subset(df, rml == 1)
df_rml <- df_rml[, col_names]

cor_matrix <- model.matrix(~0+., data=df_rml) %>% cor(use="pairwise.complete.obs")



#col_names <- c("rml", "rmlsl", "regag", "lcauth", "edib", "conc",
#               "thclim", "rtl", "pblus", "rmlhc", "estlm", 
#               "tax2", "prcng", "pkr", "warnr", "adgn", "adyth",
#               "undrg2", "drvpr2",  "opncn")

new_labels <- c(
  "Recreational Cannabis Legalization", "Recreational Cannabis Sales",
  "Regulatory Agency", "Local Authority", "Edibles Allowed", 
  "Concentrates Allowed", "Potency Limits", "Pulic Use Restrictions",
  "Retail Sales Consumption Restrictions", 
  "Home Cultivation", "Home Delivery", "Establishment Limits", 
  "Taxation Scheme", "Pricing Controls", "Packaging Requirements", 
  "Warning Labels Requirements", "Advertising Restrictions", 
  "Youth Advertising Restrictions", "Underage Restrictions", 
  "Driving Prohibitions", "Open Container Laws"
)

colnames(cor_matrix) <- new_labels
rownames(cor_matrix) <- new_labels

p_rml <- ggcorrplot(
  cor_matrix,
  show.diag = FALSE,
  type = "lower",
  lab = TRUE,
  lab_size = 2,
  # method = "circle",
  outline.color = "white"#,
  # insig = "blank",
  #title = paste("Correlation Among RML State-Years")
)

p_rml

```

# Main models

```{r}
#| echo: false
#| warning: false

cannabispol <- df[order(df$ID), ] 


row.names(cannabispol) <- NULL


col_names <- c("rml", "mml", "rmlsl", "regag", "lcauth", "edib", "hemp",  
               "adgn", "adyth","pkr", "warnr", "pkwrn", "prcng", "hmdl",
               "thclim", "rtl", "pblus", "mmlhc", "rmlhc", "tax","tax2",
               "undrg2", "drvpr2",  "drvpr3", "opncn", "opncn2", "estlm")


#Change categorical variables to factor variables
cannabispol[,col_names] <- lapply(cannabispol[, col_names], factor) 

######################################################################
#Prepare data for Mplus

#All state-years
cannabispol_subset <- cannabispol[, c("ID", "styr" ,"stabbr", "stfips", "year", col_names)]
cannabispol_subset <- cannabispol_subset %>% arrange(styr)

names(cannabispol_subset)


state <- subset(cannabispol, select = c("ID", "styr", "stname", "stabbr", "stfips", "year"))
state <- state %>% arrange(styr)



```

## Pooled latent class analysis

### Standard model

```{r, results='hide'}
#| label: Run LCA standard model 
#| echo: false
#| warning: false
#| eval: false


output_dir <- here("results", "main", "sd")
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}


# Run models 1-5 classes with all variables
lca_pooled_rml_k1_5 <- lapply(1:5, function(k) {
  lca_pooled_rml_enum <- mplusObject(
    TITLE = glue("Class {k} Standard LCA"),
    
    VARIABLE = glue(
      "categorical = lcauth thclim 
      rtl pblus hmdl tax2 prcng 
      pkwrn adgn drvpr2 opncn;
      
      usevar =  lcauth thclim 
      rtl pblus hmdl tax2 prcng 
      pkwrn adgn drvpr2 opncn;
      
      classes = c({k});
      
      idvariable = id;
     " 
    ),
    
    ANALYSIS =
      "type = mixture;
      starts = 500 100;
      lrtstarts = 0 0 1000 500;
      stiteration = 20;",
    
    
    SAVEDATA = glue("file = c{k}_lca_standard.txt;
      save = cprob;
      format = free;"),
    
    OUTPUT = "tech10 tech11 tech14;",
    
    usevariables = colnames(cannabispol_subset),
    rdata = cannabispol_subset
  )
  
  lca_pooled_rml_enum_fit <- mplusModeler(
    lca_pooled_rml_enum,
    dataout = file.path(output_dir, glue("c{k}_lca_standard.dat")),
    modelout = file.path(output_dir, glue("c{k}_lca_standard.inp")),
    check = TRUE,
    run = TRUE,
    hashfilename = FALSE
  )
})



```

#### Standard model 1-5 class comparison

```{r}
#| label: LCA standard models
#| echo: false
#| warning: false


lca_main_standard <-  readModels(here("results", "main", "sd"), quiet = TRUE)

enum_summary_main_standard <-  LatexSummaryTable(
  lca_main_standard,
  keepCols = c(
    "Title",
    "Parameters",
    "LL",
    "BIC",
    "aBIC",
    "BLRT_PValue",
    "T11_VLMR_PValue",
    "Observations"#,
    #"Entropy"
  ),
  sortBy = "Title"
)



allFit_main_standard <- enum_summary_main_standard %>% 
  mutate(
    aBIC = -2*LL + Parameters*log((Observations+2)/24),
    CAIC = -2*LL + Parameters*(log(Observations)+1),
    AWE = -2*LL + 2*Parameters*(log(Observations)+1.5),
    SIC = -0.5 * BIC,
    expSIC = exp(SIC - max(SIC)),
    BF = exp(SIC - dplyr::lead(SIC)),
    cmPk = expSIC / sum(expSIC)
  ) %>% 
  select(Title, Parameters, LL, BIC, aBIC, CAIC, AWE, BLRT_PValue, T11_VLMR_PValue, BF, cmPk) %>% 
  arrange(Parameters)



allFit_main_standard %>% 
  mutate(Title = str_remove(Title, " LCA Enumeration ")) %>% 
  gt() %>%
  tab_header(
    title = md("<b>Model Fit Summary Table: All States 2013 - 2024</b>"), 
    subtitle = md("Standard model")) %>% 
  cols_label(
    Title = "Classes",
    Parameters = md("Par"),
    LL = md("*LL*"),
    T11_VLMR_PValue = "VLMR",
    BLRT_PValue = "BLRT",
    BF = md("BF"),
    cmPk = md("*cmPk*")#,
#    Entropy = md("Entropy")
)%>%
  tab_footnote(
    footnote = md(
    "*Note.* Par = parameters; *LL* = log likelihood;
      BIC = bayesian information criterion;
      aBIC = sample size adjusted BIC; CAIC = consistent Akaike information criterion;
      AWE = approximate weight of evidence criterion;
      BLRT = bootstrapped likelihood ratio test p-value;
      VLMR = Vuong-Lo-Mendell-Rubin adjusted likelihood ratio test p-value;
      cmPk = approximate correct model probability."), 
    locations = cells_title()) %>% 
  tab_options(column_labels.font.weight = "bold") %>% 
  fmt_number(10,decimals = 2,
             drop_trailing_zeros=TRUE,
             suffixing = TRUE) %>% 
  fmt_number(c(3:9,11), 
             decimals = 0) %>% 
  sub_missing(1:11,
              missing_text = "--") %>% 
  fmt(c(8:9,11),
    fns = function(x) 
    ifelse(x<0.001, "<.001",
           scales::number(x, accuracy = 0.01))) %>%
  fmt(10, fns = function(x) 
    ifelse(x>100, ">100",
           scales::number(x, accuracy = .1))) 

```

```{r , fig.width=7, fig.height=7, out.extra="width=0.9\\linewidth"}
#| label: LCA resuls by state and year
#| echo: false
#| warning: false

lca4_sd <-  state |> inner_join(lca_main_standard$c4_lca_standard.out$savedata,
                             by = 'ID')


#Recoded
lca4_sd$lawtype_sd[lca4_sd$C == 3] <- 0 #States without RML
lca4_sd$lawtype_sd[lca4_sd$C == 1] <- 1 #Early legalization
lca4_sd$lawtype_sd[lca4_sd$C == 2] <- 2 #More regulated than 1
lca4_sd$lawtype_sd[lca4_sd$C == 4] <- 3 #More regulated than 2


table(lca4_sd$C, lca4_sd$lawtype_sd)


rmlyear <- cannabispol_subset |>
  arrange((styr)) |> 
  group_by(stabbr) |> 
  filter(rml == 1) |> 
  dplyr::slice(1) |> 
  ungroup()

rmlslyear <- cannabispol_subset |> 
  arrange((styr)) |> 
  group_by(stabbr)  |> 
  filter(rmlsl == 1)  |> 
  dplyr::slice(1) |> 
  ungroup()


cols <-  c("0" = "#CBD5E8",
           "1" = "#FDCDAC",
           "2" = "#B3E2CD",
           "3" = "#F4CAE4")


#c("#F8766D",  # reddish
#  "#7CAE00",  # green
##  "#00BFC4",  # cyan/teal
#  "#C77CFF")  # purple

cols <- c( "0" = "#B3CDE3", 
           "1" = "#FBB4AE", 
           "2" = "#CCEBC5", 
           "3" = "#DECBE4")

cols <- c("0" = "#F8766D",
          "1" = "#7CAE00",
          "2" = "#00BFC4",
          "3" = "#C77CFF")


cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

#c("0" = "#F8766D",  # reddish
#  "1" = "#7CAE00",  # green
#  "2" = "#00BFC4",  # cyan/teal
#  "3" = "#C77CFF")  # purple

lca4_sd_gp <- 
  ggplot(lca4_sd, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(lawtype_sd)
                       )
         ) +
  geom_tile(color = "gray100") +
  geom_tile(
    data = rmlyear,
    mapping = aes(x = as.factor(year), y = stabbr,
                  linetype = "RCL start"),
    #shape = 15,
    linewidth = 0.25,
    color = "black",
    fill = NA,
    inherit.aes = FALSE
    ) +
  geom_point(
    data = rmlslyear,
   mapping =  aes(x = as.factor(year), y = stabbr,
                  shape = "RCL sales start"),
    shape = 15,
    size = 1,
    color = "black",
 #   show.legend = TRUE,
   inherit.aes = FALSE
   ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
 # scale_shape_manual(
#    name = "",
#    values = c("RCL sales start" = 15),   # solid circle
#    labels = c("RCL sales start" = "RCL Sales Start")
#  ) +
  scale_linetype_manual(
    name = "",
    values = c("RCL start" = 1),   # solid line (used for rectangle outline)
    labels = c("RCL start" = "RCL Start")
  ) +
  scale_shape_manual(
    name = "",
    values = c("RCL sales start" = 15),   # solid line (used for rectangle outline)
    labels = c("RCL sales start" = "RCL sales start")
  ) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Standard LCA all states-years 2013-2024")


lca4_sd_gp

```

```{r , fig.width=7, fig.height=8, out.extra="width=0.9\\linewidth"}

# 1) a dummy point inside the panel so ggplot can draw a legend key for "RCL start"
dummy_pt <- with(lca4_sd, data.frame(
  year   = as.factor(sort(unique(year))[1]),
  stabbr = sort(unique(stabbr))[1]
))

lca4_sd_gp <-
  ggplot(lca4_sd, aes(x = as.factor(year), y = stabbr,
                      fill = as.factor(lawtype_sd))) +
  geom_tile(color = "gray100") +
  
  # draw the rectangle outline in the plot, but NO legend from this layer
  geom_tile(
    data = rmlyear,
    aes(x = as.factor(year), y = stabbr),
    linewidth = 0.25, color = "black", fill = NA,
    inherit.aes = FALSE, show.legend = FALSE
  ) +
  
  # actual sales start squares (will provide one legend entry)
  geom_point(
    data = rmlslyear,
    aes(x = as.factor(year), y = stabbr, shape = "Sales start"),
    color = "black", size = 1.2, inherit.aes = FALSE
  ) +
  
  # legend-only entry for RCL start (invisible on plot, but appears in legend)
  geom_point(
    data = dummy_pt,
    aes(x = year, y = stabbr, shape = "RCL start"),
    color = "black", size = 0, alpha = 0, inherit.aes = FALSE
  ) +
  
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("1: No RCL","2: Pre-commercial","3: Full Access","4: Dispensary Access"),
                    name   = "Latent Class") +
  
  # ONE combined legend block for both markers
  scale_shape_manual(
    name   = "",
    breaks = c("RCL start","Sales start"),
    values = c("RCL start" = 0,  # open square = law enacted
               "Sales start" = 15),  # filled square = sales start
    labels = c("RCL start"  = "RCL starts",
               "Sales start" = "RCL sales start")
  ) +
  
  guides(
    fill  = guide_legend(order = 1,
                         byrow = TRUE, 
                         keyheight = unit(10, "pt")),
    shape = guide_legend(order = 2,
                         byrow = TRUE,
                         override.aes = list(size = 3, color = "black", alpha = 1),
                         keyheight = unit(11, "pt"))
  ) +
  theme_minimal() +
  theme(
    legend.spacing.y  = unit(3, "pt"),       # no extra gap between rows
    legend.key.height = unit(10, "pt"),
    legend.box.spacing = unit(4, "pt")
  ) +
  labs(x = "Year", y = "State",
       title = "Standard LCA all states-years 2013–2024")

lca4_sd_gp

```

### Examples by state

### Nevada

```{r , fig.width=9, fig.height=1.2, out.extra="width=0.9\\linewidth"}
#| label: LCA resuls for nevada by year
#| echo: false
#| warning: false

lca4_sd_nv <-  subset(lca4_sd, stabbr == "NV" ) 
rmlyear_nv <-  subset(rmlyear, stabbr == "NV" ) 
rmlslyear_nv <- subset(rmlslyear, stabbr == "NV" ) 



cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

#c("0" = "#F8766D",  # reddish
#  "1" = "#7CAE00",  # green
#  "2" = "#00BFC4",  # cyan/teal
#  "3" = "#C77CFF")  # purple

lca4_sd_nv_gp <- 
  ggplot(lca4_sd_nv, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(lawtype_sd)
                       )
         ) +
  geom_tile(color = "gray100") +
  geom_tile(
    data = rmlyear_nv,
    mapping = aes(x = as.factor(year), y = stabbr,
                  linetype = "RCL start"),
    #shape = 15,
    linewidth = 0.25,
    color = "black",
    fill = NA,
    inherit.aes = FALSE
    ) +
  geom_point(
    data = rmlslyear_nv,
   mapping =  aes(x = as.factor(year), y = stabbr,
                  shape = "RCL sales start"),
    shape = 15,
    size = 1,
    color = "black",
 #   show.legend = TRUE,
   inherit.aes = FALSE
   ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
 # scale_shape_manual(
#    name = "",
#    values = c("RCL sales start" = 15),   # solid circle
#    labels = c("RCL sales start" = "RCL Sales Start")
#  ) +
  scale_linetype_manual(
    name = "",
    values = c("RCL start" = 1),   # solid line (used for rectangle outline)
    labels = c("RCL start" = "RCL Start")
  ) +
  scale_shape_manual(
    name = "",
    values = c("RCL sales start" = 15),   # solid line (used for rectangle outline)
    labels = c("RCL sales start" = "RCL sales start")
  ) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Standard LCA all states-years 2013-2024")


lca4_sd_nv_gp

```

### Colorado

```{r , fig.width=9, fig.height=1.2, out.extra="width=0.9\\linewidth"}
#| label: LCA resuls for Colorado and year
#| echo: false
#| warning: false

lca4_sd_co <-  subset(lca4_sd, stabbr == "CO" ) 
rmlyear_co <-  subset(rmlyear, stabbr == "CO" ) 
rmlslyear_co <- subset(rmlslyear, stabbr == "CO" ) 



cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

#c("0" = "#F8766D",  # reddish
#  "1" = "#7CAE00",  # green
#  "2" = "#00BFC4",  # cyan/teal
#  "3" = "#C77CFF")  # purple

lca4_sd_co_gp <- 
  ggplot(lca4_sd_co, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(lawtype_sd)
                       )
         ) +
  geom_tile(color = "gray100") +
  geom_tile(
    data = rmlyear_co,
    mapping = aes(x = as.factor(year), y = stabbr,
                  linetype = "RCL start"),
    #shape = 15,
    linewidth = 0.25,
    color = "black",
    fill = NA,
    inherit.aes = FALSE
    ) +
  geom_point(
    data = rmlslyear_co,
   mapping =  aes(x = as.factor(year), y = stabbr,
                  shape = "RCL sales start"),
    shape = 15,
    size = 1,
    color = "black",
 #   show.legend = TRUE,
   inherit.aes = FALSE
   ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
 # scale_shape_manual(
#    name = "",
#    values = c("RCL sales start" = 15),   # solid circle
#    labels = c("RCL sales start" = "RCL Sales Start")
#  ) +
  scale_linetype_manual(
    name = "",
    values = c("RCL start" = 1),   # solid line (used for rectangle outline)
    labels = c("RCL start" = "RCL Start")
  ) +
  scale_shape_manual(
    name = "",
    values = c("RCL sales start" = 15),   # solid line (used for rectangle outline)
    labels = c("RCL sales start" = "RCL sales start")
  ) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Standard LCA all states-years 2013-2024")


lca4_sd_co_gp



```

### Maine

```{r , fig.width=9, fig.height=1.2, out.extra="width=0.9\\linewidth"}
#| label: LCA resuls for Maine by year
#| echo: false
#| warning: false

lca4_sd_me <-  subset(lca4_sd, stabbr == "ME" ) 
rmlyear_me <-  subset(rmlyear, stabbr == "ME" ) 
rmlslyear_me <- subset(rmlslyear, stabbr == "ME" ) 



cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

#c("0" = "#F8766D",  # reddish
#  "1" = "#7CAE00",  # green
#  "2" = "#00BFC4",  # cyan/teal
#  "3" = "#C77CFF")  # purple

lca4_sd_me_gp <- 
  ggplot(lca4_sd_me, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(lawtype_sd)
                       )
         ) +
  geom_tile(color = "gray100") +
  geom_tile(
    data = rmlyear_me,
    mapping = aes(x = as.factor(year), y = stabbr,
                  linetype = "RCL start"),
    #shape = 15,
    linewidth = 0.25,
    color = "black",
    fill = NA,
    inherit.aes = FALSE
    ) +
  geom_point(
    data = rmlslyear_me,
   mapping =  aes(x = as.factor(year), y = stabbr,
                  shape = "RCL sales start"),
    shape = 15,
    size = 1,
    color = "black",
 #   show.legend = TRUE,
   inherit.aes = FALSE
   ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
 # scale_shape_manual(
#    name = "",
#    values = c("RCL sales start" = 15),   # solid circle
#    labels = c("RCL sales start" = "RCL Sales Start")
#  ) +
  scale_linetype_manual(
    name = "",
    values = c("RCL start" = 1),   # solid line (used for rectangle outline)
    labels = c("RCL start" = "RCL Start")
  ) +
  scale_shape_manual(
    name = "",
    values = c("RCL sales start" = 15),   # solid line (used for rectangle outline)
    labels = c("RCL sales start" = "RCL sales start")
  ) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Standard LCA all states-years 2013-2024")


lca4_sd_me_gp

```

#### Item endorsement probability by class

```{r}

probabilities <- lca_main_standard$c4_lca_standard.out$parameters$probability.scale

itemprobability <- probabilities %>%
  mutate(
    # map 3→1, 1→2, 2→3, 4→4 by releveling then relabeling
    LatentClassRecoded = factor(LatentClass, levels = c(3, 1, 2, 4), labels = 1:4)
  ) %>%
  select(param, category, est, LatentClassRecoded) %>%
  pivot_wider(
    names_from  = LatentClassRecoded,
    values_from = est,
    names_prefix = "Class ",
    names_expand = TRUE, names_sort = TRUE
  ) %>%
  select(param, category, `Class 1`, `Class 2`, `Class 3`, `Class 4`)


print(itemprobability, n = 52)

```

#### Item endorsement probability graphs

```{r}
#| label: Prepare lables and data for radar graph
#| warning: false

probabilities$itemcat <- paste(probabilities$param, probabilities$category)


#label_map <- c(
#  "ADGN 2"   = "Advertising - P/C",
#  "ADGN 3"   = "Advertising - Both",
#  "DRVPR2 2" = "Driving - No specific",
#  "DRVPR2 3" = "Driving - No THC",
#  "DRVPR2 4" = "Driving - THC lim",
#  "HMDL 2"   = "Home Delivery - Allowed",
#  "HMDL 3"   = "Home Delivery - Prohibited",
#  "LCAUTH 2" = "Local opt - Control only",
#  "LCAUTH 3" = "Local opt - Both",
#  "OPNCN 2"  = "Open containers in vehicles no law",
#  "OPNCN 3"  = "Open containers in vehicles prohibited",
#  "PBLUS 2"  = "Public Use - Prohibited",
#  "PBLUS 3"  = "Public Use - Restricted",
#  "PKWRN 2"  = "Packaging or warning required",
#  "PRCNG 2"  = "Pricing controls",
#  "RTL 2"    = "Retail use - Off-prem",
#  "RTL 3"    = "Retail - Off + On",
#  "TAX2 2"  = "Tax - Sales only",
#  "TAX2 3"    = "Tax - Weight only",
#  "TAX2 4"    = "Tax - Sales + Weight",
#  "TAX2 5"    = "Tax - Potency",
#  "THCLIM 2"    = "Potency limits"
#)

# Create lables for the categories

label_map <- c(
  "LCAUTH 2" = "Local opt - Ctrl",
  "LCAUTH 3" = "Local opt - Both",
  "THCLIM 2" = "THC lim",
  "RTL 2"    = "Rtl - Off",
  "RTL 3"    = "Rtl - Off+On",
  "PBLUS 2"  = "Public use - Restr",
  "PBLUS 3"  = "Public use - Proh",
  "HMDL 2"   = "Delivery - Restr",
  "HMDL 3"   = "Delivery - Proh",
  "TAX2 2"   = "Tax - Sales",
  "TAX2 3"   = "Tax - Weight",
  "TAX2 4"   = "Tax - S+W",
  "TAX2 5"   = "Tax - Potency",
  "PRCNG 2"  = "Pricing - Ctrl",
  "PKWRN 2"  = "Pack/warn - Req",
  "ADGN 2"   = "Mkt - P/C",
  "ADGN 3"   = "Mkt - Both",
  "DRVPR2 2" = "Driving - None",
  "DRVPR2 3" = "Driving - No THC",
  "DRVPR2 4" = "Driving - THC lim",
  "OPNCN 2"  = "Open cont - No law",
  "OPNCN 3"  = "Open cont - Proh"
)



#Remove the reference category
probabilities2 <- subset(probabilities, category !=1)
probabilities2 <- subset(probabilities2, itemcat  != "DRVPR2 2" )
probabilities2 <- subset(probabilities2, itemcat  != "OPNCN 2" )


#Recoded
probabilities2$LatentClass_recoded[probabilities2$LatentClass == 3] <- 1 #States without RML
probabilities2$LatentClass_recoded[probabilities2$LatentClass == 1] <- 2 #Early legalization
probabilities2$LatentClass_recoded[probabilities2$LatentClass == 2] <- 3 #More regulated than 1
probabilities2$LatentClass_recoded[probabilities2$LatentClass == 4] <- 4 #More regulated than 2




```

##### Plot for latent class 2

```{r, fig.width=8, fig.height=5, out.extra="width=0.9\\linewidth"}
#| label: radar plot for LCA1
#| warning: false

# Latent class 2
lca2_plot_df <- subset(probabilities2, LatentClass_recoded==2)


plot_df <- lca2_plot_df |> 
  mutate(itemcat = factor(itemcat)) |> 
  group_by(itemcat) |> 
  mutate(sum_est_param = sum(est, na.rm = TRUE)) |>   # for ordering
  ungroup()

plot_df <- plot_df %>%
  mutate(
    itemcat_lab = dplyr::recode(itemcat, !!!label_map),
    itemcat_lab = factor(
      itemcat_lab,
      levels = c(
        "Local opt - Ctrl",
        "Local opt - Both",
        "THC lim",
        "Rtl - Off",
        "Rtl - Off+On",
        "Public use - Restr",
        "Public use - Proh",
        "Delivery - Restr",
        "Delivery - Proh",
        "Tax - Sales",
        "Tax - Weight",
        "Tax - S+W",
        "Tax - Potency",
        "Pricing - Ctrl",
        "Pack/warn - Req",
        "Mkt - P/C",
        "Mkt - Both",
        "Driving - None",
        "Driving - No THC",
        "Driving - THC lim",
        "Open cont - No law",
        "Open cont - Proh"
      )
    )
  )

max_r <- max(plot_df$est, na.rm = TRUE)
max_r <- ceiling(max_r * 1.1 * 10) / 10               # pad & round to 0.1s
grid_breaks <- seq(0, max_r, length.out = 4)  

lt2 <- ggplot(plot_df) +
  # custom radial gridlines
  geom_hline(
    aes(yintercept = y),
    data = data.frame(y = seq(0, max(plot_df$est, na.rm = TRUE), length.out = 4)),
    color = "lightgrey", linewidth = 0.3
  ) +

  # bars: est by itemcat for each param
  geom_col(
    aes(
      x = str_wrap(itemcat_lab, 6),
      y = est,
      fill = itemcat
    ),
    position = position_dodge2(width = 0.9, preserve = "single"),
    alpha = 0.9,
    width = 0.85
  ) +
  # Annotate custom scale inside plot
  #annotate(
  #  x = 11.7, 
   # y = 0.5, 
  #  label = "0.5", 
  #  geom = "text", 
  #  color = "gray12",
  #  size = 3
#    family = "Bell MT"
  #) +
  #annotate(
  ##  x = 11.7, 
  #  y =1, 
  #  label = "1", 
  #  geom = "text", 
  #  color = "gray12",
  #  size = 3
#    family = "Bell MT"
  #) +

  coord_polar() +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 8),
    legend.position = "none"   # remove legend
  )

lt2 + labs(title = "Pre-commercial model")
```

```{r}


library(tibble)

key_df <- tribble(
  ~label,                ~description,
  "Local opt - Ctrl",    "Local control authorized",
  "Local opt - Both",    "Local opt-out and control authorized",
  "THC lim",             "THC limits apply",
  "Rtl - Off",           "Off-premises sales allowed only",
  "Rtl - Off+On",        "On- and off-premises sales allowed",
  "Public use - Restr",  "Public use restricted",
  "Public use - Proh",   "Public use prohibited",
  "Delivery - Restr",    "Home delivery allowed",
  "Delivery - Proh",     "Home delivery prohibited",
  "Tax - Sales",         "Sales-based taxes",
  "Tax - Weight",        "Weight-based taxes",
  "Tax - S+W",           "Sales + weight-based taxes",
  "Tax - Potency",       "Potency-based taxes",
  "Pricing - Ctrl",      "Price controls in place",
  "Pack/warn - Req",     "THC & serving size on label required",
  "Mkt - P/C",           "Ad Placement or content restricted",
  "Mkt - Both",          "Ad Placement and content restricted",
  "Driving - None",      "General driving prohibitions",
  "Driving - No THC",    "No THC limits, evidence required",
  "Driving - THC lim",   "THC limits specified",
  "Open cont - No law",  "No law on open containers",
  "Open cont - Proh",    "Open containers prohibited"
)



```

```{r}

plot_df <- plot_df %>%
  left_join(key_df, by = c("itemcat_lab" = "label"))

caption_text <- paste0(
  key_df$label, ": ", key_df$description, collapse = "; "
)

lt2 +
  labs(caption = caption_text) +
  theme(
    plot.caption = element_text(hjust = 0, size = 7, color = "gray30")
  )

```

```{r, fig.width=8, fig.height=5, out.extra="width=0.9\\linewidth"}
#| label: radar plot for LCA3
#| warning: false

# Latent class 3
lca3_plot_df <- subset(probabilities2, LatentClass_recoded==3)


plot_df <- lca3_plot_df |> 
  mutate(itemcat = factor(itemcat)) |> 
  group_by(itemcat) |> 
  mutate(sum_est_param = sum(est, na.rm = TRUE)) |>   # for ordering
  ungroup()

#plot_df <- plot_df %>%
#  mutate(itemcat_lab = recode(itemcat, !!!label_map))

plot_df <- plot_df %>%
  mutate(
    itemcat_lab = dplyr::recode(itemcat, !!!label_map),
    itemcat_lab = factor(
      itemcat_lab,
      levels = c(
        "Local opt - Ctrl",
        "Local opt - Both",
        "THC lim",
        "Rtl - Off",
        "Rtl - Off+On",
        "Public use - Restr",
        "Public use - Proh",
        "Delivery - Restr",
        "Delivery - Proh",
        "Tax - Sales",
        "Tax - Weight",
        "Tax - S+W",
        "Tax - Potency",
        "Pricing - Ctrl",
        "Pack/warn - Req",
        "Mkt - P/C",
        "Mkt - Both",
        "Driving - None",
        "Driving - No THC",
        "Driving - THC lim",
        "Open cont - No law",
        "Open cont - Proh"
      )
    )
  )

max_r <- max(plot_df$est, na.rm = TRUE)
max_r <- ceiling(max_r * 1.1 * 10) / 10               # pad & round to 0.1s
grid_breaks <- seq(0, max_r, length.out = 4)  

lt3 <- ggplot(plot_df) +
  # custom radial gridlines
  geom_hline(
    aes(yintercept = y),
    data = data.frame(y = seq(0, max(plot_df$est, na.rm = TRUE), length.out = 4)),
    color = "lightgrey", linewidth = 0.3
  ) +

  # bars: est by itemcat for each param
  geom_col(
    aes(
      x = str_wrap(itemcat_lab, 6),
      y = est,
      fill = itemcat
    ),
    position = position_dodge2(width = 0.9, preserve = "single"),
    alpha = 0.9,
    width = 0.85
  ) +

  # optional: a faint radial guide per param (turn on if you like)
  # geom_segment(
  #   aes(
  #     x = reorder(itemcat, sum_est_param),
  #     xend = reorder(itemcat, sum_est_param),
  #     y = 0, yend = max_r
  #   ),
  #   linetype = "dashed", color = "gray30", linewidth = 0.2
  # ) +

  coord_polar() +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 8),
    legend.position = "none"   # remove legend
  )

lt3 + labs(title = "Full access model")

```

```{r, fig.width=8, fig.height=5, out.extra="width=0.9\\linewidth"}
#| label: radar plot for LCA4
#| warning: false

# Latent class 3
lca4_plot_df <- subset(probabilities2, LatentClass_recoded==4)


plot_df <- lca4_plot_df |> 
  mutate(itemcat = factor(itemcat)) |> 
  group_by(itemcat) |> 
  mutate(sum_est_param = sum(est, na.rm = TRUE)) |>   # for ordering
  ungroup()

#plot_df <- plot_df %>%
#  mutate(itemcat_lab = recode(itemcat, !!!label_map))

plot_df <- plot_df %>%
  mutate(
    itemcat_lab = dplyr::recode(itemcat, !!!label_map),
    itemcat_lab = factor(
      itemcat_lab,
      levels = c(
        "Local opt - Ctrl",
        "Local opt - Both",
        "THC lim",
        "Rtl - Off",
        "Rtl - Off+On",
        "Public use - Proh",
        "Public use - Restr",
        "Delivery - Restr",
        "Delivery - Proh",
        "Tax - Sales",
        "Tax - Weight",
        "Tax - S+W",
        "Tax - Potency",
        "Pricing - Ctrl",
        "Pack/warn - Req",
        "Mkt - P/C",
        "Mkt - Both",
        "Driving - None",
        "Driving - No THC",
        "Driving - THC lim",
        "Open cont - No law",
        "Open cont - Proh"
      )
    )
  )

max_r <- max(plot_df$est, na.rm = TRUE)
max_r <- ceiling(max_r * 1.1 * 10) / 10               # pad & round to 0.1s
grid_breaks <- seq(0, max_r, length.out = 4)  

lt4 <- ggplot(plot_df) +
  # custom radial gridlines
  geom_hline(
    aes(yintercept = y),
    data = data.frame(y = seq(0, max(plot_df$est, na.rm = TRUE), length.out = 4)),
    color = "lightgrey", linewidth = 0.3
  ) +

  # bars: est by itemcat for each param
  geom_col(
    aes(
      x = str_wrap(itemcat_lab, 6),
      y = est,
      fill = itemcat
    ),
    position = position_dodge2(width = 0.9, preserve = "single"),
    alpha = 0.9,
    width = 0.85
  ) +

  # optional: a faint radial guide per param (turn on if you like)
  # geom_segment(
  #   aes(
  #     x = reorder(itemcat, sum_est_param),
  #     xend = reorder(itemcat, sum_est_param),
  #     y = 0, yend = max_r
  #   ),
  #   linetype = "dashed", color = "gray30", linewidth = 0.2
  # ) +

  coord_polar() +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 8),
    legend.position = "none"   # remove legend
  )

lt4 + labs(title = "Dispensary Access Model")
```

# Supplement material

## Multilevel results

```{r}
#| label: LCA multilevel models
#| echo: false
#| warning: false


lca_main_multilevel <-  readModels(here("results", "s2", "ml"), quiet = TRUE)


enum_summary_main_multilevel <-  LatexSummaryTable(
  lca_main_multilevel,
  keepCols = c(
    "Title",
    "Parameters",
    "LL",
    "BIC",
    "aBIC",
    #"BLRT_PValue",
    "T11_VLMR_PValue",
    "Observations"#,
    #"Entropy"
  ),
  sortBy = "Title"
)



allFit_main_multilevel <- enum_summary_main_multilevel %>% 
  mutate(
    aBIC = -2*LL + Parameters*log((Observations+2)/24),
    CAIC = -2*LL + Parameters*(log(Observations)+1),
    AWE = -2*LL + 2*Parameters*(log(Observations)+1.5),
    SIC = -0.5 * BIC,
    expSIC = exp(SIC - max(SIC)),
    BF = exp(SIC - dplyr::lead(SIC)),
    cmPk = expSIC / sum(expSIC)
  ) %>% 
  select(Title, Parameters, LL, BIC, aBIC, CAIC, AWE, #BLRT_PValue, 
         T11_VLMR_PValue, BF, cmPk
         ) %>% 
  arrange(Parameters)



allFit_main_multilevel %>% 
  mutate(Title = str_remove(Title, " LCA Enumeration ")) %>% 
  gt() %>%
  tab_header(
    title = md("<b>Model Fit Summary Table: All States 2013 - 2024</b>"), 
    subtitle = md("Multilevel model")) %>% 
  cols_label(
    Title = "Classes",
    Parameters = md("Par"),
    LL = md("*LL*"),
    T11_VLMR_PValue = "VLMR"#,
   # BLRT_PValue = "BLRT",
    #BF = md("BF"),
  #  cmPk = md("*cmPk*")#,
#    Entropy = md("Entropy")
)%>%
  tab_footnote(
    footnote = md(
    "*Note.* Par = parameters; *LL* = log likelihood;
      BIC = bayesian information criterion;
      aBIC = sample size adjusted BIC; CAIC = consistent Akaike information criterion;
      AWE = approximate weight of evidence criterion;
      VLMR = Vuong-Lo-Mendell-Rubin adjusted likelihood ratio test p-value."), 
    locations = cells_title()) %>% 
  tab_options(column_labels.font.weight = "bold") %>% 
  fmt_number(10,decimals = 2,
             drop_trailing_zeros=TRUE,
             suffixing = TRUE) %>% 
  fmt_number(c(3:9,10), 
             decimals = 0) %>% 
  sub_missing(1:10,
              missing_text = "--") %>% 
  fmt(c(8:9,10),
    fns = function(x) 
    ifelse(x<0.001, "<.001",
           scales::number(x, accuracy = 0.01))) %>%
  fmt(10, fns = function(x) 
    ifelse(x>100, ">100",
           scales::number(x, accuracy = .1))) 



```

```{r , fig.width=7, fig.height=7, out.extra="width=0.9\\linewidth"}
#| label: LCA multilevel models plot recoded
#| echo: false
#| warning: false

lca4_ml <-  state |> inner_join(lca_main_multilevel$c4_lca_multilevel.out$savedata,
                             by = 'ID')


#Recoded
lca4_ml$lawtype_ml[lca4_ml$C == 1] <- 0 #States that havent legalized
lca4_ml$lawtype_ml[lca4_ml$C == 3] <- 1 #Early legalization
lca4_ml$lawtype_ml[lca4_ml$C == 4] <- 2 #More regulated than 1
lca4_ml$lawtype_ml[lca4_ml$C == 2] <- 3 #More regulated than 2


table(lca4_ml$C, lca4_ml$lawtype_ml)


rmlyear <- cannabispol_subset |>
  arrange((styr)) |> 
  group_by(stabbr) |> 
  filter(rml == 1) |> 
  dplyr::slice(1) |> 
  ungroup()

rmlslyear <- cannabispol_subset |> 
  arrange((styr)) |> 
  group_by(stabbr)  |> 
  filter(rmlsl == 1)  |> 
  dplyr::slice(1) |> 
  ungroup()


cols <-  c("0" = "#CBD5E8",
           "1" = "#FDCDAC",
           "2" = "#B3E2CD",
           "3" = "#F4CAE4")


#c("#F8766D",  # reddish
#  "#7CAE00",  # green
##  "#00BFC4",  # cyan/teal
#  "#C77CFF")  # purple

cols <- c( "0" = "#B3CDE3", 
           "1" = "#FBB4AE", 
           "2" = "#CCEBC5", 
           "3" = "#DECBE4")

cols <- c("0" = "#F8766D",
          "1" = "#7CAE00",
          "2" = "#00BFC4",
          "3" = "#C77CFF")


cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

#c("0" = "#F8766D",  # reddish
#  "1" = "#7CAE00",  # green
#  "2" = "#00BFC4",  # cyan/teal
#  "3" = "#C77CFF")  # purple

lca4_ml_gp <- 
  ggplot(lca4_ml, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(lawtype_ml)
                       )
         ) +
  geom_tile(color = "gray100") +
  geom_tile(
    data = rmlyear,
    mapping = aes(x = as.factor(year), y = stabbr,
                  linetype = "RCL start"),
    #shape = 15,
    linewidth = 0.25,
    color = "black",
    fill = NA,
    inherit.aes = FALSE
    ) +
  geom_point(
    data = rmlslyear,
   mapping =  aes(x = as.factor(year), y = stabbr,
                  shape = "RCL sales start"),
    shape = 15,
    size = 1,
    color = "black",
 #   show.legend = TRUE,
   inherit.aes = FALSE
   ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
 # scale_shape_manual(
#    name = "",
#    values = c("RCL sales start" = 15),   # solid circle
#    labels = c("RCL sales start" = "RCL Sales Start")
#  ) +
  scale_linetype_manual(
    name = "",
    values = c("RCL start" = 1),   # solid line (used for rectangle outline)
    labels = c("RCL start" = "RCL Start")
  ) +
  scale_shape_manual(
    name = "",
    values = c("RCL sales start" = 15),   # solid line (used for rectangle outline)
    labels = c("RCL sales start" = "RCL sales start")
  ) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Multilevel LCA all states-years 2013-2024")


lca4_ml_gp
```

```{r , fig.width=7, fig.height=7, out.extra="width=0.9\\linewidth"}
#| label: LCA multilevel models plot with labels
#| echo: false
#| warning: false

#library(grid)
# 1) a dummy point inside the panel so ggplot can draw a legend key for "RCL start"
dummy_pt <- with(lca4_ml, data.frame(
  year   = as.factor(sort(unique(year))[1]),
  stabbr = sort(unique(stabbr))[1]
))

lca4_ml_gp <-
  ggplot(lca4_ml, aes(x = as.factor(year), y = stabbr,
                      fill = as.factor(lawtype_ml))) +
  geom_tile(color = "gray100") +
  
  # draw the rectangle outline in the plot, but NO legend from this layer
  geom_tile(
    data = rmlyear,
    aes(x = as.factor(year), y = stabbr),
    linewidth = 0.25, color = "black", fill = NA,
    inherit.aes = FALSE, show.legend = FALSE
  ) +
  
  # actual sales start squares (will provide one legend entry)
  geom_point(
    data = rmlslyear,
    aes(x = as.factor(year), y = stabbr, shape = "Sales start"),
    color = "black", size = 1.2, inherit.aes = FALSE
  ) +
  
  # legend-only entry for RCL start (invisible on plot, but appears in legend)
  geom_point(
    data = dummy_pt,
    aes(x = year, y = stabbr, shape = "RCL start"),
    color = "black", size = 0, alpha = 0, inherit.aes = FALSE
  ) +
  
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1","Class 2","Class 3","Class 4"),
                    name   = "Latent Class") +
  
  # ONE combined legend block for both markers
  scale_shape_manual(
    name   = "",
    breaks = c("RCL start","Sales start"),
    values = c("RCL start" = 0,  # open square = law enacted
               "Sales start" = 15),  # filled square = sales start
    labels = c("RCL start"  = "RCL starts",
               "Sales start" = "RCL sales start")
  ) +
  
  guides(
    fill  = guide_legend(order = 1,
                         byrow = TRUE, 
                         keyheight = unit(10, "pt")),
    shape = guide_legend(order = 2,
                         byrow = TRUE,
                         override.aes = list(size = 3, color = "black", alpha = 1),
                         keyheight = unit(11, "pt"))
  ) +
  theme_minimal() +
  theme(
    legend.spacing.y  = unit(3, "pt"),       # no extra gap between rows
    legend.key.height = unit(10, "pt"),
    legend.box.spacing = unit(4, "pt")
  ) +
  labs(x = "Year", y = "State",
       title = "Multilevel LCA all states-years 2013–2024")

lca4_ml_gp

```

```{r}
#| label: probabilities_ml

#Recoded
#lca4_ml$lawtype_ml[lca4_ml$C == 1] <- 0 #States that havent legalized
#lca4_ml$lawtype_ml[lca4_ml$C == 3] <- 1 #Early legalization
#lca4_ml$lawtype_ml[lca4_ml$C == 4] <- 2 #More regulated than 1
#lca4_ml$lawtype_ml[lca4_ml$C == 2] <- 3 #More regulated than 2

probabilities_ml <- lca_main_multilevel$c4_lca_multilevel.out$parameters$probability.scale


probabilities_ml$LatentClass_recoded[probabilities_ml$LatentClass  == 1] <- 1 #States that havent legalized
probabilities_ml$LatentClass_recoded[probabilities_ml$LatentClass  == 4] <- 3 #More regulated than 2
probabilities_ml$LatentClass_recoded[probabilities_ml$LatentClass  == 3] <- 2 #Early legalization
probabilities_ml$LatentClass_recoded[probabilities_ml$LatentClass  == 2] <- 4 #More regulated than 3



probabilities_ml <- probabilities_ml |>
  select(param, category, est, LatentClass_recoded) |>
  pivot_wider(names_from = LatentClass_recoded,
              values_from = est,
              names_prefix = "Class")


print(probabilities_ml, n = 52)

```

## Results without DC

```{r , fig.width=7, fig.height=7, out.extra="width=0.9\\linewidth"}
#| label: LCA multilevel without DC models plot
#| echo: false
#| warning: false


lca_noDC_multilevel <-  readModels(here("results", "noDC", "ml"), quiet = TRUE)


state_noDC  <- subset(state, stabbr != "DC")

lca4_noDC_ml <-  state_noDC |> inner_join(lca_noDC_multilevel$savedata,
                             by = 'ID')

probabilities_noDC_ml <- lca_noDC_multilevel$parameters$probability.scale

# probabilities_noDC_ml <- probabilities_noDC_ml |>
#   select(param, category, est, LatentClass ) |>
#   pivot_wider(names_from = LatentClass ,
#               values_from = est,
#               names_prefix = "Class")


itemprobability_noDC_ml <- probabilities_noDC_ml %>%
  mutate(
    # map 3→1, 1→2, 2→3, 4→4 by releveling then relabeling
    LatentClassRecoded = factor(LatentClass, levels = c(3, 4, 1, 2), labels = 1:4)
  ) %>%
  select(param, category, est, LatentClassRecoded) %>%
  pivot_wider(
    names_from  = LatentClassRecoded,
    values_from = est,
    names_prefix = "Class ",
    names_expand = TRUE, names_sort = TRUE
  ) %>%
  select(param, category, `Class 1`, `Class 2`, `Class 3`, `Class 4`)


print(itemprobability_noDC_ml, n = 52)


#Recoded
lca4_noDC_ml$lawtype_ml[lca4_noDC_ml$C == 3] <- 0 #States that havent legalized
lca4_noDC_ml$lawtype_ml[lca4_noDC_ml$C == 4] <- 1 #Early legalization
lca4_noDC_ml$lawtype_ml[lca4_noDC_ml$C == 1] <- 2 #More regulated than 1
lca4_noDC_ml$lawtype_ml[lca4_noDC_ml$C == 2] <- 3 #More regulated than 2


table(lca4_ml$C, lca4_ml$lawtype_ml)


rmlyear <- cannabispol_subset |>
  arrange((styr)) |> 
  group_by(stabbr) |> 
  filter(rml == 1) |> 
  dplyr::slice(1) |> 
  ungroup()

rmlyear <- subset(rmlyear, stabbr!="DC")

rmlslyear <- cannabispol_subset |> 
  arrange((styr)) |> 
  group_by(stabbr)  |> 
  filter(rmlsl == 1)  |> 
  dplyr::slice(1) |> 
  ungroup()

# cols <-  c("0" = "#CBD5E8",
#            "1" = "#FDCDAC",
#            "2" = "#B3E2CD",
#            "3" = "#F4CAE4")


#c("#F8766D",  # reddish
#  "#7CAE00",  # green
##  "#00BFC4",  # cyan/teal
#  "#C77CFF")  # purple

# cols <- c( "0" = "#B3CDE3", 
#            "1" = "#FBB4AE", 
#            "2" = "#CCEBC5", 
#            "3" = "#DECBE4")
# 
# cols <- c("0" = "#F8766D",
#           "1" = "#7CAE00",
#           "2" = "#00BFC4",
#           "3" = "#C77CFF")


cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

#c("0" = "#F8766D",  # reddish
#  "1" = "#7CAE00",  # green
#  "2" = "#00BFC4",  # cyan/teal
#  "3" = "#C77CFF")  # purple

lca4_no_DC_ml_gp <- 
  ggplot(lca4_noDC_ml, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(lawtype_ml)
                       )
         ) +
  geom_tile(color = "gray100") +
  geom_tile(
    data = rmlyear,
    mapping = aes(x = as.factor(year), y = stabbr,
                  linetype = "RCL start"),
    #shape = 15,
    linewidth = 0.25,
    color = "black",
    fill = NA,
    inherit.aes = FALSE
    ) +
  geom_point(
    data = rmlslyear,
   mapping =  aes(x = as.factor(year), y = stabbr,
                  shape = "RCL sales start"),
    shape = 15,
    size = 1,
    color = "black",
 #   show.legend = TRUE,
   inherit.aes = FALSE
   ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
 # scale_shape_manual(
#    name = "",
#    values = c("RCL sales start" = 15),   # solid circle
#    labels = c("RCL sales start" = "RCL Sales Start")
#  ) +
   # ONE combined legend block for both markers
  scale_shape_manual(
    name   = "",
    breaks = c("RCL start","Sales start"),
    values = c("RCL start" = 0,  # open square = law enacted
               "Sales start" = 15),  # filled square = sales start
    labels = c("RCL start"  = "RCL starts",
               "Sales start" = "RCL sales start")
  ) +
  
  # scale_linetype_manual(
  #   name = "",
  #   values = c("RCL start" = 1),   # solid line (used for rectangle outline)
  #   labels = c("RCL start" = "RCL Start")
  # ) +
  # scale_shape_manual(
  #   name = "",
  #   values = c("RCL sales start" = 15),   # solid line (used for rectangle outline)
  #   labels = c("RCL sales start" = "RCL sales start")
  # ) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Multilevel LCA all states-years excluding DC 2013-2024")


lca4_no_DC_ml_gp
```

```{r, fig.width=7, fig.height=7, out.extra="width=0.9\\linewidth"}
#| label: standard without DC models plot
#| echo: false
#| warning: false


#lca_noDC_multilevel <-  readModels(here("results", "noDC", "ml"), quiet = TRUE)


lca_noDC_standard <-  readModels(here("results", "noDC", "sd"), quiet = TRUE)

state_noDC  <- subset(state, stabbr != "DC")

results_noDC <-  state |> inner_join(lca_noDC_standard$c4_lca_standard.out$savedata, by = 'ID')


t#able(results_noDC$C)


probabilities_noDC_sd <- lca_noDC_standard$c4_lca_standard.out$parameters$probability.scale



itemprobability_noDC_sd <- probabilities_noDC_sd %>%
  mutate(
    # map 3→1, 1→2, 2→3, 4→4 by releveling then relabeling
    LatentClassRecoded = factor(LatentClass, levels = c(1, 2, 3, 4), labels = 1:4)
  ) %>%
  select(param, category, est, LatentClassRecoded) %>%
  pivot_wider(
    names_from  = LatentClassRecoded,
    values_from = est,
    names_prefix = "Class ",
    names_expand = TRUE, names_sort = TRUE
  ) %>%
  select(param, category, `Class 1`, `Class 2`, `Class 3`, `Class 4`)


print(itemprobability_noDC_sd, n = 52)


#Recoded
results_noDC$lawtype_sd[results_noDC$C == 3] <- 0 #States that havent legalized
results_noDC$lawtype_sd[results_noDC$C == 1] <- 1 #Pre-commercial
results_noDC$lawtype_sd[results_noDC$C == 2] <- 2 #Full Access
results_noDC$lawtype_sd[results_noDC$C == 4] <- 3 #Dispensary Access


table(results_noDC$C, results_noDC$lawtype_sd)


rmlyear <- cannabispol_subset |>
  arrange((styr)) |> 
  group_by(stabbr) |> 
  filter(rml == 1) |> 
  dplyr::slice(1) |> 
  ungroup()

rmlyear <- subset(rmlyear, stabbr!="DC")

rmlslyear <- cannabispol_subset |> 
  arrange((styr)) |> 
  group_by(stabbr)  |> 
  filter(rmlsl == 1)  |> 
  dplyr::slice(1) |> 
  ungroup()

# cols <-  c("0" = "#CBD5E8",
#            "1" = "#FDCDAC",
#            "2" = "#B3E2CD",
#            "3" = "#F4CAE4")


#c("#F8766D",  # reddish
#  "#7CAE00",  # green
##  "#00BFC4",  # cyan/teal
#  "#C77CFF")  # purple

# cols <- c( "0" = "#B3CDE3", 
#            "1" = "#FBB4AE", 
#            "2" = "#CCEBC5", 
#            "3" = "#DECBE4")
# 
# cols <- c("0" = "#F8766D",
#           "1" = "#7CAE00",
#           "2" = "#00BFC4",
#           "3" = "#C77CFF")


cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

#c("0" = "#F8766D",  # reddish
#  "1" = "#7CAE00",  # green
#  "2" = "#00BFC4",  # cyan/teal
#  "3" = "#C77CFF")  # purple

lca4_no_DC_sd_gp <- 
  ggplot(results_noDC, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(lawtype_sd)
                       )
         ) +
  geom_tile(color = "gray100") +
  geom_tile(
    data = rmlyear,
    mapping = aes(x = as.factor(year), y = stabbr,
                  linetype = "RCL start"),
    #shape = 15,
    linewidth = 0.25,
    color = "black",
    fill = NA,
    inherit.aes = FALSE
    ) +
  geom_point(
    data = rmlslyear,
   mapping =  aes(x = as.factor(year), y = stabbr,
                  shape = "RCL sales start"),
    shape = 15,
    size = 1,
    color = "black",
 #   show.legend = TRUE,
   inherit.aes = FALSE
   ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
 # scale_shape_manual(
#    name = "",
#    values = c("RCL sales start" = 15),   # solid circle
#    labels = c("RCL sales start" = "RCL Sales Start")
#  ) +
   # ONE combined legend block for both markers
  scale_shape_manual(
    name   = "",
    breaks = c("RCL start","Sales start"),
    values = c("RCL start" = 0,  # open square = law enacted
               "Sales start" = 15),  # filled square = sales start
    labels = c("RCL start"  = "RCL starts",
               "Sales start" = "RCL sales start")
  ) +
  
  # scale_linetype_manual(
  #   name = "",
  #   values = c("RCL start" = 1),   # solid line (used for rectangle outline)
  #   labels = c("RCL start" = "RCL Start")
  # ) +
  # scale_shape_manual(
  #   name = "",
  #   values = c("RCL sales start" = 15),   # solid line (used for rectangle outline)
  #   labels = c("RCL sales start" = "RCL sales start")
  # ) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Standard LCA all states-years excluding DC 2013-2024")


lca4_no_DC_sd_gp
```

## Results 2020-2024

### Standard

```{r, fig.width=5, fig.height=8, out.extra="width=0.9\\linewidth"}
#| label: LCA standard models plot 2020 2024
#| echo: false
#| warning: false


lca_2020_2024_sd <-  readModels(here("results", "2020_2024", "sd"), quiet = TRUE)


state_2020_2024 <- subset(state, year>2019)

lca4_2020_2024_state <-  state_2020_2024 |> inner_join(lca_2020_2024_sd$savedata, by = 'ID')

#Recoded
lca4_2020_2024_state$lawtype_sd[lca4_2020_2024_state$C == 4] <- 0 #States that havent legalized
lca4_2020_2024_state$lawtype_sd[lca4_2020_2024_state$C == 3] <- 1 #Early legalization
lca4_2020_2024_state$lawtype_sd[lca4_2020_2024_state$C == 1] <- 2 #More regulated than 1
lca4_2020_2024_state$lawtype_sd[lca4_2020_2024_state$C == 2] <- 3 #More regulated than 2

cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

lca4_sd_2020_2024_gp <- 
  ggplot(lca4_2020_2024_state, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(lawtype_sd)
                       )
         ) +
  geom_tile(color = "gray100") +
 # geom_tile(
#    data = rsdyear,
#    mapping = aes(x = as.factor(year), y = stabbr),
#    #shape = 15,
#    linewidth = 0.25,
#    color = "black",
#    fill = NA,
#    inherit.aes = FALSE
#    ) +
#  geom_point(
#    data = rsdslyear,
#   mapping =  aes(x = as.factor(year), y = stabbr),
#    shape = 15,
#    size = 1,
#    color = "black",
#    show.legend = FALSE,
#   inherit.aes = FALSE
 #  ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Standard LCA all states-years 2020-2024")


lca4_sd_2020_2024_gp

```

### Multilevel

```{r, fig.width=5, fig.height=8, out.extra="width=0.9\\linewidth"}
#| label: LCA multilevel models plot 2020 2024
#| echo: false
#| warning: false


lca_2020_2024_ml <-  readModels(here("results", "2020_2024", "ml"), quiet = TRUE)


state_2020_2024 <- subset(state, year>2019)

lca4_2020_2024_state <-  state_2020_2024 |> inner_join(lca_2020_2024_ml$savedata, by = 'ID')

#Recoded
lca4_2020_2024_state$lawtype_ml[lca4_2020_2024_state$C == 4] <- 0 #States that havent legalized
lca4_2020_2024_state$lawtype_ml[lca4_2020_2024_state$C == 3] <- 1 #Early legalization
lca4_2020_2024_state$lawtype_ml[lca4_2020_2024_state$C == 2] <- 2 #More regulated than 1
lca4_2020_2024_state$lawtype_ml[lca4_2020_2024_state$C == 1] <- 3 #More regulated than 2


cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

lca4_ml_2020_2024_gp <- 
  ggplot(lca4_2020_2024_state, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(lawtype_ml)
                       )
         ) +
  geom_tile(color = "gray100") +
 # geom_tile(
#    data = rmlyear,
#    mapping = aes(x = as.factor(year), y = stabbr),
#    #shape = 15,
#    linewidth = 0.25,
#    color = "black",
#    fill = NA,
#    inherit.aes = FALSE
#    ) +
#  geom_point(
#    data = rmlslyear,
#   mapping =  aes(x = as.factor(year), y = stabbr),
#    shape = 15,
#    size = 1,
#    color = "black",
#    show.legend = FALSE,
#   inherit.aes = FALSE
 #  ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Multilevel LCA all states-years 2020-2024")


lca4_ml_2020_2024_gp

```

## Results in RCL states

```{r , fig.width=7, fig.height=5, out.extra="width=0.9\\linewidth"}


results_rcl_all <-  readModels(here("results", "rcl"), quiet = TRUE)


results_RCL <- results_rcl_all$c4_lca_rcl.out$savedata


results_RCL <-  state |> inner_join(results_RCL,
                                             by = 'ID')


results_RCL$C_recoded[results_RCL$C == 3] <- 0 #States without RML
results_RCL$C_recoded[results_RCL$C == 4] <- 1 #Early legalization
results_RCL$C_recoded[results_RCL$C == 1] <- 2 #More regulated than 1
results_RCL$C_recoded[results_RCL$C == 2] <- 3 #More regulated than 2

###########

cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")


lca4_rcl_gp <- 
  ggplot(results_RCL, 
         mapping = aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(C_recoded)
         )
  ) +
  geom_tile(color = "gray100") +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
  # scale_shape_manual(
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Standard LCA in RCL states 2013-2024")


lca4_rcl_gp
```

## Results by year

```{r, fig.width=5, fig.height=8, out.extra="width=0.9\\linewidth"}

results_yr <-  readModels(here("results", "yr"), quiet = TRUE)





results_yr_2020 <- results_yr$c4_lca_2020.out$savedata

results_yr_2020$year

#Recoded

results_yr_2021 <- results_yr$c4_lca_2021.out$savedata
results_yr_2022 <- results_yr$c4_lca_2022.out$savedata
results_yr_2023 <- results_yr$c4_lca_2023.out$savedata
results_yr_2024 <- results_yr$c4_lca_2024.out$savedata

results_yr_2024$C_recoded[results_yr_2024$C == 4] <- 0 #States without RML
results_yr_2024$C_recoded[results_yr_2024$C == 2] <- 1 #Early legalization
results_yr_2024$C_recoded[results_yr_2024$C == 3] <- 2 #More regulated than 1
results_yr_2024$C_recoded[results_yr_2024$C == 1] <- 3 #More regulated than 2

results_yr_2023$C_recoded[results_yr_2023$C == 4] <- 0 #States without RML
results_yr_2023$C_recoded[results_yr_2023$C == 3] <- 1 #Early legalization
results_yr_2023$C_recoded[results_yr_2023$C == 2] <- 2 #More regulated than 1
results_yr_2023$C_recoded[results_yr_2023$C == 1] <- 3 #More regulated than 2

results_yr_2022$C_recoded[results_yr_2022$C == 2] <- 0 #States without RML
results_yr_2022$C_recoded[results_yr_2022$C == 4] <- 1 #Early legalization
results_yr_2022$C_recoded[results_yr_2022$C == 3] <- 2 #More regulated than 1
results_yr_2022$C_recoded[results_yr_2022$C == 1] <- 3 #More regulated than 2

results_yr_2021$C_recoded[results_yr_2021$C == 3] <- 0 #States without RML
results_yr_2021$C_recoded[results_yr_2021$C == 4] <- 1 #Early legalization
results_yr_2021$C_recoded[results_yr_2021$C == 2] <- 2 #More regulated than 1
results_yr_2021$C_recoded[results_yr_2021$C == 1] <- 3 #More regulated than 2

results_yr_2020$C_recoded[results_yr_2020$C == 4] <- 0 #States without RML
results_yr_2020$C_recoded[results_yr_2020$C == 1] <- 1 #Early legalization
results_yr_2020$C_recoded[results_yr_2020$C == 2] <- 2 #More regulated than 1
results_yr_2020$C_recoded[results_yr_2020$C == 3] <- 3 #More regulated than 2


results_yr_2020_2024 <- rbind(results_yr_2020, results_yr_2021, results_yr_2022,
                              results_yr_2023, results_yr_2024)

results_yr_2020_2024 <-  state |> inner_join(results_yr_2020_2024,
                                by = 'ID')


cols <- c("0" = "#dddddd",
          "1" = "#FFB000",
          "2" = "#19c691",
          "3" = "#ba91ff")

#results_yr_2020$C_recoded <- as.factor(results_yr_2020$C_recoded)

lca_yr_2020_2024_gp <- 
  ggplot(results_yr_2020_2024, 
         mapping =  aes(x = as.factor(year),
                       y = stabbr,
                       fill = as.factor(C_recoded)
         )
  ) +
  geom_tile(color = "gray100") +
  # geom_tile(
  #   data = rmlyear,
  #   mapping = aes(x = as.factor(year), y = stabbr,
  #                 linetype = "RCL start"),
  #   #shape = 15,
  #   linewidth = 0.25,
  #   color = "black",
  #   fill = NA,
  #   inherit.aes = FALSE
  # ) +
  # geom_point(
  #   data = rmlslyear,
  #   mapping =  aes(x = as.factor(year), y = stabbr,
  #                  shape = "RCL sales start"),
  #   shape = 15,
  #   size = 1,
  #   color = "black",
  #   #   show.legend = TRUE,
  #   inherit.aes = FALSE
  # ) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = cols,
                    labels = c("Class 1", "Class 2", "Class 3", "Class 4")) +
  # scale_shape_manual(
  #    name = "",
  #    values = c("RCL sales start" = 15),   # solid circle
  #    labels = c("RCL sales start" = "RCL Sales Start")
  #  ) +
  # scale_linetype_manual(
  #   name = "",
  #   values = c("RCL start" = 1),   # solid line (used for rectangle outline)
  #   labels = c("RCL start" = "RCL Start")
  # ) +
  # scale_shape_manual(
  #   name = "",
  #   values = c("RCL sales start" = 15),   # solid line (used for rectangle outline)
  #   labels = c("RCL sales start" = "RCL sales start")
  # ) +
  theme_minimal() +
  labs(x = "Year",
       y = "State",
       fill = "Latent Class",
       title = "Standard LCA all states-years 2020-2024")


lca_yr_2020_2024_gp


plot_df <- results_yr_2020_2024 |>
  dplyr::mutate(
    year = factor(year),
    C_recoded = factor(C_recoded, levels = c(1,2,3,4))  # or as.character if your levels are "1"-"4"
  )


```
